---
author: Piotr Wójcik
title: Measuring inequalities from space. Analysis of satellite raster images with R
output:
  html_document:
    theme: spacelab
    highlight: tango
    number_sections: true
    toc: true
    toc_float: true 
---
Let's install needed packages.

```{r, eval=FALSE}
install.packages(c("rgdal", "rgeos", "dplyr", 
                   "raster", "tabularaster",
                   "ggplot2", "ggrepel", 
                   "digest", "maptools", "readr",
                   "sf", "eurostat", "WDI"))
```

```{r, warning=FALSE, message=FALSE}
library(rgdal)
library(rgeos)
library(dplyr)
library(raster)
library(tabularaster)
library(ggplot2)
library(ggrepel)
library(digest)
library(maptools)
library(readr)
library(sf)
library(eurostat)
library(WDI)
```

The first detailed example will be explained for European countries. 

Similar analyses for European regions, US states and world countries will follow.

# Step 1. importing raster data and map into R

## Raster data

In this tutorial I used DMSP-OLS data for 2013 zipped in [this archive](https://ngdc.noaa.gov/eog/data/web_data/v4composites/F182013.v4.tar) -- a specific file is called `F182013.v4c_web.stable_lights.avg_vis.tif` (it's size - 692 MB).
Due to GitHub file size limitations you have to download this file manually and store in the local `files` folder.

Raster data is commonly used to represent spatially continuous phenomena. A raster divides the world into a grid of equally sized rectangles (referred to as cells or, in the context of satellite remote sensing, pixels) that all have one or more values (or missing values) for the variables of interest.

The `raster()` function is used to import data in the `raster` package, if the file has only one layer (i.e. for each cell / pixel there is only one value stored in it) or the `brick()` function, if there are more layers.

Nighttime light data contains only one layer

```{r}
localfile <- "files/F182013.v4c_web.stable_lights.avg_vis.tif"

data_raster <- raster(localfile)
```

## Shapefile for European countries

I will limit the raster data only to the EU area. To do this, let's also load shapefile file with a map of EU countries (obtained from [Eurostat](https://ec.europa.eu/eurostat/web/gisco/geodata/reference-data/administrative-units-statistical-units) - a particular file used is `UTS_RG_10M_2016_4326_LEVL_0.shp.zip`, NUTS0 is the national level).

```{r, message=FALSE, warning=FALSE}
map_nuts0 <- st_read("files/NUTS_RG_10M_2013_4326_LEVL_0.shp")
plot(st_geometry(map_nuts0))
```

Let's limit the map to the mainland (we cut all the islands).
For spatial limitation of `sf` objects one can use a function `st_crop()`, in which the second argument is the object defining the bounding box.

```{r}
# check the current bounding box
st_bbox(map_nuts0)

# define own bounding box
my_bbox = c(xmin = -11, 
            ymin = 30, 
            xmax = 55.83616, 
            ymax = 71.15304)

# limit the map
map_nuts0 <- st_crop(map_nuts0,
                     my_bbox)
```

Lets see the map again.

```{r}
plot(st_geometry(map_nuts0))
```

One can also use `ggplot2` for mapping `sf` objects.

```{r}
ggplot(map_nuts0) +
  geom_sf() +
  theme_classic()
```

Presenting the values of the selected variable requires mapping its name on the `fill` aestetics.

```{r}
ggplot(map_nuts0) + 
  geom_sf(aes(fill = CNTR_CODE)) +
  theme_classic()
```


# Step 2. checking and adjusting

To be able to impose a map on raster data, one has to make sure that THE SAME geographical coordinate projection is used in BOTH datasets.

Let's check what are the current values of these parameters.

```{r}
crs(data_raster)

crs(map_nuts0)
```

Seems identical (`WGS84`).

If there is a need to unify the projections, one can simply copy the one used in the raster file into the map. The `st_transform()` function can be used for this:

```{r}
map_nuts0 <- st_transform(map_nuts0,
                          crs(data_raster)@projargs)
```


# Step 3. limitation of the extent of raster data

To apply spatial limitation of the raster data set one has to use a function `crop()`, in which the second argument is the extent of the area. It must be an `Extent` class object, resulting from the use of the `extent()` function.

Lets see the extent of our raster data and for the map:

```{r}  
extent(data_raster)

extent(map_nuts0)
```

We crop the raster data to the rectangle, where the European Union is located.

```{r}
data_raster_EU <- crop(data_raster,      # source data
                       extent(map_nuts0)) # extent
```

Let's see the result.

```{r}
plot(data_raster_EU)
```

One can clearly see European continent and major cities. Lets super-impose the map of EU regions.

```{r}
plot(data_raster_EU)
plot(st_geometry(map_nuts0),
     add = TRUE) # added on previous map
```

This data still contains areas outside the EU. Raster data for pixels outside the polygons can be reemoved (set to missing) with the `mask()` function.

```{r}
data_raster_EU <- mask(data_raster_EU, 
                       mask = map_nuts0)
```

The more polygons (regions), the more time it takes (here ca. 40 seconds).

Let's display data truncated to the EU area imposing once again the grid of countries borders.

```{r}
plot(data_raster_EU)
plot(st_geometry(map_nuts0), 
     add = TRUE)
```

Data have been correctly restricted to the European Union area.

# Step 4. aggregation to spatial units

Very efficient identification and aggregation of data for selected areas can be done by using the `cellnumbers()` function from the [`tabularaster` package by Michael D. Sumner](https://cran.r-project.org/web/packages/tabularaster/vignettes/tabularaster-usage.html). It allows to identify numbers (indexes) of cells located in every region (polygon in the map object). And aggregation based on indexed data is ca. 500 times faster than with using purely the `extract()` function.

The only problem here is that some of the countries have several islands (the geometry of type MULTIPOLYGON) which is not YET SUPPORTED by `cellnumbers()`.

So for this simple example lets convert all geometries to just `POLYGON`s (inc case of a `MULTIPOLYGON` only the first will be used).

```{r, message=FALSE, warning=FALSE}
map_nuts0 <- st_cast(map_nuts0, "POLYGON")
```

And finally the aggregation.

```{r}
# identify numbers (indexes) of cells located
# in every region (polygon in the map object)
cell <- cellnumbers(data_raster_EU,
                    map_nuts0)
# which creates a data.frame with 
# two columns: object_ and cell_

# and then aggregate the values 
# for all cell_(s)  by object_

cell %>% 
  mutate(light = raster::extract(data_raster_EU, 
                                 cell$cell_)) %>% 
  group_by(object_) %>% 
  summarise(lights = sum(light, na.rm = TRUE)) -> lights_NUTS0
```

Lets add a new column to the map data frame.

```{r}
map_nuts0$lights2013 <- lights_NUTS0$lights
```

and plot it on a map 

```{r}
ggplot(map_nuts0) + 
  geom_sf(aes(fill = lights2013)) +
  coord_sf() +
  scale_fill_gradient(low = "black", high = "yellow") +
  guides(fill = FALSE) +
  theme_classic()
```

# Correlations with statistical data 

Lets in addtion import data on GDP and population from Eurostat.

```{r}
eurostat <- read_csv("files/Eurostat_data.csv")
```

To merge them correctly I convert all values to character.

```{r}
map_nuts0$NUTS_ID <- as.character(map_nuts0$NUTS_ID)
eurostat$NUTS_ID <- as.character(eurostat$NUTS_ID)

map_nuts0_2 <- 
  map_nuts0 %>% 
  left_join(eurostat, by = "NUTS_ID")
```

Lets check correlations between lights, log(lights) and GDP, GDP per capita and population.

```{r}
map_nuts0_2 %>% 
  mutate(l_lights2013 = log(lights2013)) %>% 
  dplyr::select(population2013, gdp2013, gdpcap2013, lights2013, l_lights2013) %>% 
  st_drop_geometry() %>% 
  cor(use = "pairwise.complete.obs") 
```

Scatterplot between GDP and lights.

```{r}
ggplot(map_nuts0_2,
       aes(x = gdp2013/1e+06,
           y = lights2013/1e+06)) +
  xlab("GDP in 2013") +
  ylab("NTLI in 2013") +
 geom_point(size = 5, 
             col = "darkblue") +
  geom_smooth(method = "lm", se = FALSE) +
  geom_label_repel(data = map_nuts0_2 %>% 
                     filter(gdp2013/1e+06 > 1 | lights2013/1e+06 > 3),
                   aes(label = NUTS_ID),
                   size = 5) +
  geom_text(x = 2, y = 3,
            size = 8,
            label = "cor(GDP, NTLI) = 0.88") +
  geom_text(x = 2, y = 2,
            size = 8,
            label = "cor(GDP, lnNTLI) = 0.63") +
  # title
  ggtitle("Night-time lights intensity vs GDP in 2013 in EU countries") +
  theme_bw()
```  

Scatterplot between population and lights.

```{r}
ggplot(map_nuts0_2,
       aes(x = population2013/1e+04,
           y = lights2013/1e+06)) +
  xlab("population in 2013") +
  ylab("NTLI in 2013") +
  geom_point(size = 5, 
             col = "darkblue") +
  geom_smooth(method = "lm", se = FALSE) +
  geom_label_repel(data = map_nuts0_2 %>%
                     filter(population2013/1e+04 > 3 |
                              lights2013/1e+06 > 3),
                   aes(label = NUTS_ID),
                   size = 5) +
  geom_text(x = 5.5, y = 3,
            size = 8,
            label = "cor(pop, NTLI) = 0.88") +
  geom_text(x = 5.5, y = 2,
            size = 8,
            label = "cor(pop, lnNTLI) = 0.67") +
  ggtitle("Night-time lights intensity vs population in 2013 in EU countries") +
  theme_bw()
```

# Analysis for European NUTS 2 regions

## Importing shapefile

Import shapefile for European NUTS2 regions and limit its area to the mainland of Europe (cut the islands as before).

```{r, message=FALSE, warning=FALSE}
map_nuts2 <- st_read("files/NUTS_RG_10M_2013_4326_LEVL_2.shp")
```

```{r, error=TRUE}
map_nuts2 <- st_crop(map_nuts2,
                     my_bbox)
```
          
## Checking validity

To be able to limit the raster data to the desired area one has to make sure that should check if all geometries in the `sf` object are valid. Intuitively the two-dimensional geometry (`POLYGON`) is valid, if its boundaries are created from subsequent **connecting**, but **non-overlapping** and **non-intersecting** sections. `MULTIPOLYGON` geometry is valid if it consists of valid POLYGONs.

Checking the validity of the geometries of the analyzed objects can be time-consuming (especially for very complex geometries), but it saves problems in their subsequent visualization or analysis.

```{r}
table(st_is_valid(map_nuts2))
```

Nine regions have incorrect geometries. They might me automatically corrected with `st_make_valid()` function from `lwgeom` package, which is the extension of `sf`. Its description can be found here: [github.com/r-spatial/lwgeom](https://github.com/r-spatial/lwgeom).

```{r}
map_nuts2 <- lwgeom::st_make_valid(map_nuts2)
```

And apply the `crop()` function once again.

```{r}
map_nuts2 <- st_crop(map_nuts2,
                     my_bbox)
```                

Now it works fine.
Next we adjust the coordinates.

```{r}
map_nuts2 <- st_transform(map_nuts2,
                          crs(data_raster_EU)@projargs)
```

And plot the map of EU NUTS2 regions superimposed on night-time lights distribution.

```{r}
plot(data_raster_EU)
plot(st_geometry(map_nuts2),
     add = TRUE) # added on previous map
```

## Aggregation

For the use of `cellnumbers()` we need to simplify complex geometries to just POLYGONs.

```{r, message=FALSE, warning=FALSE}
map_nuts2 <- st_cast(map_nuts2, "POLYGON")
```

And apply aggregation.

```{r}
cell <- cellnumbers(data_raster_EU,
                    map_nuts2)
cell %>% 
  mutate(light = raster::extract(data_raster_EU, 
                                 cell$cell_)) %>% 
  group_by(object_) %>% 
  summarise(lights = sum(light, na.rm = TRUE)) -> lights_nuts2
```

Lets add a new column to the map data frame.

```{r, error=TRUE}
map_nuts2$lights2013 <- lights_nuts2$lights
```

Lights data is missing for one region - which?

```{r}
(which_missing <- which(!1:311 %in% lights_nuts2$object_))

map_nuts2[which_missing,]
plot(map_nuts2[which_missing,"CNTR_CODE"])
```

This region has a very strange shape -- for simplicity, let's assume it has the intensity of night lights equal to 0.

```{r}
lights_nuts2 <- lights_nuts2 %>% 
  rbind(., data.frame(object_ = which_missing,
                      lights = 0)) %>% 
  arrange(object_)
```

Add a new column to the map data frame.

```{r}
map_nuts2$lights2013 <- lights_nuts2$lights
```

And finally plot it on a map

```{r}
ggplot(map_nuts2) +
  geom_sf(aes(fill = lights2013)) +
  coord_sf() +
  scale_fill_gradient(low = "black", high = "yellow") +
  guides(fill = FALSE) +
  theme_classic()
```

## Importing data from Eurostat

Lets in addtion import data on GDP and population for regions directly from Eurostat - with the use of `eurostat` package -- see [package tutorial](https://ropengov.github.io/eurostat/articles/eurostat_tutorial.html).

To download data, you need to know their codes used by Eurostat.

This can be checked on the [Eurostat website](https://ec.europa.eu/eurostat/data/database) or in particular reports, for example [GDP at regional level](http://ec.europa.eu/eurostat/statistics-explained/index.php/GDP_at_regional_level), [Population statistics at regional level](http://ec.europa.eu/eurostat/statistics-explained/index.php/Population_statistics_at_regional_level).

You can also check them by searching the Eurostat database by keywords:

```{r}
results <- search_eurostat("GDP")
head(results[, 1:3])
```

Let's download the GDP figures in the NUTS2 regions (`nama_10r_2gdp`).

```{r}
gdp <- get_eurostat("nama_10r_2gdp", 
                    time_format = "num")

# filter just global GDP in PPS for 2013
gdp <- gdp %>%
  filter(time == 2013,
         unit == "MIO_PPS")
```

Similarly for population figures.

```{r}
pop <- get_eurostat("nama_10r_3popgdp", 
                    time_format = "num")

pop <- pop %>%
  filter(time == 2013)
```

We need to add this data to the `sf` object with geometries - all datasets use common Eurostat codes for regions, but columns have different names.

```{r}
map_nuts2 <- map_nuts2 %>% 
  left_join(gdp %>% dplyr::select(geo, values), 
            by = c("NUTS_ID" = "geo")) %>% 
  dplyr::rename(gdp2013 = values) %>% 
  left_join(pop %>% dplyr::select(geo, values), 
            by = c("NUTS_ID" = "geo")) %>% 
  dplyr::rename(pop2013 = values)
```

## Correlations

Finally we check for correlations.

```{r}
map_nuts2 %>% 
  mutate(l_lights2013 = log(lights2013 + 1)) %>%
  # bo był jeden region z zerem 
  dplyr::select(ends_with("2013")) %>% 
  st_drop_geometry() %>% 
  cor(use = "pairwise.complete.obs")
```

And make a scatterplot between GDP and lights.

```{r}
ggplot(map_nuts2,
       aes(x = gdp2013/100000,
           y = lights2013/100000)) +
 geom_point(size = 5, 
             colour = "darkblue") +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("NTLI in 2013") +
  geom_text(x = 2.5, y = 17,
            size = 8,
            label = "cor(GDP, NTLI) = 0.35") +
  geom_text(x = 2.5, y = 15,
            size = 8,
            label = "cor(GDP, lnNTLI) = 0.23") +
  geom_label_repel(data = map_nuts2 %>% 
                     filter(gdp2013/100000 > 2 | lights2013/100000 > 10),
                   aes(label = NUTS_NAME),
                   size = 5) +
  ggtitle("Night-time lights intensity vs GDP in 2013 in EU regions (NUTS 2)") +
  theme_bw()
```

Scatterplot between lights and population in NUTS2 regions.

```{r}
ggplot(map_nuts2,
       aes(x = pop2013/1000,
           y = lights2013/100000)) +
  geom_point(size = 5, 
             colour = "darkblue") +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("population in 2013") + ylab("lNTLI in 2013") +
  geom_text(x = 10, y = 17,
            size = 8,
            label = "cor(pop, NTLI) = 0.42") +
  geom_text(x = 10, y = 15,
            size = 8,
            label = "cor(pop, lnNTLI) = 0.66") +
  geom_label_repel(data = map_nuts2 %>% 
                     filter(pop2013/1000 > 7 |
                              lights2013/100000 > 10),
                   aes(label = NUTS_NAME),
                   size = 5) +
  ggtitle("Night-time lights intensity vs population in 2013 in EU regions (NUTS 2)") +
  theme_bw()
```

# Analysis for NUTS 3 regions

## Importing shapefile

```{r, message=FALSE, warning=FALSE}
map_nuts3 <- st_read("files/NUTS_RG_10M_2013_4326_LEVL_3.shp")
```

## Validation and adjustment

```{r}
map_nuts3 <- lwgeom::st_make_valid(map_nuts3)

map_nuts3 <- st_crop(map_nuts3,
                     my_bbox)

map_nuts3 <- st_transform(map_nuts3,
                          crs(data_raster_EU)@projargs)
```

Map of EU NUTS3 regions imposed on night-time lights intensity.

```{r}
plot(data_raster_EU)
plot(st_geometry(map_nuts3),
     add = TRUE) # added on previous map
```

## Aggregation

```{r, message=FALSE, warning=FALSE}
map_nuts3 <- st_cast(map_nuts3, "POLYGON")

cell <- cellnumbers(data_raster_EU,
                    map_nuts3)

cell %>% 
  mutate(light = raster::extract(data_raster_EU, 
                                 cell$cell_)) %>%
      group_by(object_) %>% 
  summarise(lights = sum(light, 
                         na.rm = TRUE)) -> lights_nuts3

map_nuts3$lights2013 <- lights_nuts3$lights
```

Visualization on a map 

```{r}
ggplot(map_nuts3) +
  geom_sf(aes(fill = lights2013)) +
  coord_sf() +
  scale_fill_gradient(low = "black", high = "yellow") +
  guides(fill = FALSE) +
  theme_classic()
```

## Data from Eurostat for NUTS3 regions

```{r}
gdp <- get_eurostat("nama_10r_3gdp", 
                    time_format = "num")

gdp <- gdp %>%
  filter(time == 2013,
         unit == "MIO_PPS")

pop <- get_eurostat("nama_10r_3popgdp", 
                    time_format = "num")

pop <- pop %>%
  filter(time == 2013)
```

Merging the dada with the `sf` object.

```{r}
map_nuts3 <- map_nuts3 %>% 
  left_join(gdp %>% dplyr::select(geo, values), 
            by = c("NUTS_ID" = "geo")) %>% 
  dplyr::rename(gdp2013 = values) %>% 
  left_join(pop %>% dplyr::select(geo, values), 
            by = c("NUTS_ID" = "geo")) %>% 
  dplyr::rename(pop2013 = values)
```

And checking correlations.

```{r}
map_nuts3 %>% 
  mutate(l_lights2013 = log(lights2013 + 1)) %>%
  # bo był jeden region z zerem 
  dplyr::select(ends_with("2013")) %>% 
  st_drop_geometry() %>% 
  cor(use = "pairwise.complete.obs")
```

Scatterplot between lights and GDP

```{r}
ggplot(map_nuts3,
       aes(x = gdp2013/10000,
           y = lights2013/100000)) +
  # points 
  geom_point(size = 5, 
             colour = "darkblue") +
  # regression line added
  geom_smooth(method = "lm", se = FALSE) +
  xlab("GDP in 2013") + ylab("lNTLI in 2013") +
  geom_text(x = 17, y = 6,
            size = 8,
            label = "cor(GDP, NTLI) = 0.36") +
  geom_text(x = 17, y = 5.5,
            size = 8,
            label = "cor(GDP, lnNTLI) = 0.28") +
  geom_label_repel(data = map_nuts3 %>% 
                     filter(gdp2013/10000 > 10 | lights2013/100000 > 4),
                   aes(label = NUTS_NAME),
                   size = 5) +
  # title
  ggtitle("Night-time lights intensity vs GDP in 2013 in EU regions (NUTS 3)") +
  theme_bw()
```

Scatterplot between population and nighttime lights intensity.

```{r}
ggplot(map_nuts3,
       aes(x = pop2013/1000,
           y = lights2013/100000)) +
  geom_point(size = 5, 
             colour = "darkblue") +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("population in 2013") + ylab("lNTLI in 2013") +
  geom_text(x = 4, y = 6,
            size = 8,
            label = "cor(pop, NTLI) = 0.51") +
  geom_text(x = 4, y = 5.5,
            size = 8,
            label = "cor(pop, lnNTLI) = 0.46") +
  geom_label_repel(data = map_nuts3 %>% 
                     filter(pop2013/1000 > 2 | lights2013/100000 > 3.8),
                   aes(label = NUTS_NAME),
                   size = 5) +
  ggtitle("Night-time lights intensity vs population in 2013 in EU regions (NUTS 3)") +
  theme_bw()
```

# Analysis for US states

## Import shapefile

```{r, message=FALSE, warning=FALSE}
map_US <- st_read("files/tl_2018_us_state.shp")

plot(st_geometry(map_US))
```

## Adjusting 

Adjusting coordinates projections between shapefile and raster data. 

```{r}
map_US <- st_transform(map_US, 
                       crs(data_raster)@projargs)
```

Checking validity.

```{r}
table(st_is_valid(map_US))
```

Limiting the bounding box.

```{r}
my_bbox = c(xmin = -125,
            ymin = 25, 
            xmax = -65, 
            ymax = 50)

map_US2 <- st_crop(map_US,
                   my_bbox)

plot(st_geometry(map_US2))
```

## Limiting raster data to the area of US

Then raster image is cropped to the are of US.

```{r}
data_raster_US <- crop(data_raster,    # source data
                       extent(map_US2)) # extent
```

```{r}
plot(data_raster_US)
plot(st_geometry(map_US2),
     add = TRUE)
```

One can clearly identify major US cities.

Masking the data for areas outside USA. 

```{r}
data_raster_US <- mask(data_raster_US, 
                       mask = map_US2)
```

And display data truncated to the US area imposing once again the grid of states borders.

```{r}
plot(data_raster_US)
plot(st_geometry(map_US2), 
     add = TRUE)
```

## Aggregation of lights to states

```{r, message=FALSE, warning=FALSE}
map_US2 <- st_cast(map_US2, "POLYGON")

cell <- cellnumbers(data_raster_US,
                    map_US2)
cell %>% 
  mutate(light = raster::extract(data_raster_US, 
                                 cell$cell_)) %>%
    group_by(object_) %>% 
  summarise(lights = sum(light, 
                         na.rm = TRUE)) -> lights_US

map_US2$lights2013 <- lights_US$lights
```

Result on a map. 

```{r}
ggplot(map_US2) + 
  geom_sf(aes(fill = lights2013)) +
  coord_sf() +
  scale_fill_gradient(low = "black", high = "yellow") +
  guides(fill = FALSE) +
  theme_classic()

```

## Correlation with census data

Importing data about state GDP and population.

```{r}
census <- read_csv("files/US_data_GDP_pop.csv")
```

To merge the data correctly let's convert all state codes in a map file to character values.

```{r}
map_US2$code <- as.character(map_US2$STUSPS)

map_US2_2 <- 
  map_US2 %>% 
  left_join(census, by = "code")
```

Correlation matrix 

```{r}
map_US2_2 %>% 
  mutate(l_lights2013 = log(lights2013)) %>% 
  dplyr::select(lights2013, l_lights2013, gdp2013, pop2013) %>% 
  st_drop_geometry() %>% 
  cor(use = "pairwise.complete.obs")
```

Scatterplot between GDP and lights in US states.

```{r}
ggplot(map_US2_2,
       aes(x = gdp2013/1e+5,
           y = lights2013/1e+5)) +
  xlab("GDP in 2013") +
  ylab("lNTLI in 2013") +
  geom_point(size = 5, 
             col = "darkblue") +
  geom_smooth(method = "lm", se = FALSE) +
  geom_label_repel(data = map_US2_2 %>% 
                     filter(gdp2013/1e+5 > 10),
                     aes(label = code),
                   size = 5) +
  geom_text(x = 15, y = 15,
            size = 8,
            label = "cor(GDP, NTLI) = 0.78") +
  geom_text(x = 15, y = 10,
            size = 8,
            label = "cor(GDP, lnNTLI) = 0.82") +
  ggtitle("Night-time lights intensity vs GDP in 2013 in US states") +
  theme_bw()
```

Scatterplot between population and lights in US states.

```{r}
ggplot(map_US2_2,
       aes(x = pop2013/1e+5,
           y = lights2013/1e+5)) +
  xlab("population in 2013") +
  ylab("lNTLI in 2013") +
  geom_point(size = 5, 
             col = "darkblue") +
  geom_smooth(method = "lm", se = FALSE) +
  geom_label_repel(data = map_US2_2 %>% 
                     filter(pop2013/1e+5 > 150),
                   aes(label = code),
                   size = 5) +
  geom_text(x = 250, y = 15,
            size = 8,
            label = "cor(pop, NTLI) = 0.83") +
  geom_text(x = 250, y = 10,
            size = 8,
            label = "cor(pop, lnNTLI) = 0.63") +
  ggtitle("Night-time lights intensity vs population in 2013 in US states") +
  theme_bw()
```

# Analysis for world countries

## Importing map data

In case of world counties lets use `wrld_simpl` object from the `maptools` package converted to `sf` object.

```{r}
data(wrld_simpl)
map_world <- st_as_sf(wrld_simpl)

plot(st_geometry(map_world))
```

## Checking and adjusting 

Lets adjust the coordinates projections

```{r}
crs(map_world)
crs(data_raster)

map_world <- st_transform(map_world, 
                       crs(data_raster)@projargs)
```

and check validity of geometries.

```{r}
table(st_is_valid(map_world))

map_world <- lwgeom::st_make_valid(map_world)
```

Let's also convert the type of geometries to `POLYGONs`.

```{r}
map_world2 <- st_cast(map_world, "POLYGON")
```

It appears that the extent of the map is larger than the extent of the raster data.

```{r}
extent(map_world)

(ext_r <- extent(data_raster))
```

Let's limit the map to the extent of raster object.

```{r}
my_bbox = c(xmin = ext_r@xmin, 
            ymin = ext_r@ymin, 
            xmax = ext_r@xmax, 
            ymax = ext_r@ymax)

map_world2 <- st_crop(map_world2,
                      my_bbox)
```

## Data aggregation

As operating on the whole raster data may be too demanding (in this case R will try to read all data into memory), we will make the aggregation using a loop and sum night-time lights intensity for every country separately (in fact every `POLYGON` as we have just converted original data into `POLYGONs`).

> CAUTION! This loop run for about 45 minutes on my laptop. Let me know if you find a more (time) efficient way :)

Let's collect the calculated values in a numeric vector.

```{r}
lights_world <- array(NA, nrow(map_world2))

for (i in 1:nrow(map_world2)) {
  message(i)
  map_i <- map_world2[i,]
  
  data_raster_i <- crop(data_raster,
                         extent(map_i))
  
  cell <- cellnumbers(data_raster_i,
                      map_i)
  
  if (nrow(cell) > 0) {
  cell %>% 
    mutate(light = raster::extract(data_raster_i,
                                   cell$cell_)) %>% 
    group_by(object_) %>% 
    summarise(lights = sum(light, na.rm = TRUE)) -> lights_i

    lights_world[i] <- as.numeric(lights_i$lights) 
    } else
    lights_world[i] <- 0
    
  rm(map_i, data_raster_i, cell, lights_i)
  gc()
}
```

Let's add the values to the map data and finally aggregate over `POLYGONs` within countries.

```{r}
map_world2$lights2013 <- lights_world

lights_world_counties <- map_world2 %>% 
  group_by(FIPS, ISO2, ISO3, UN, NAME) %>% 
  summarise(lights2013 = sum(lights2013, 
                             na.rm = TRUE)) %>% 
  ungroup()
```

Finally NTLI data by countries are added to the original map data.

```{r}
map_world <- map_world %>% 
  left_join(st_drop_geometry(lights_world_counties) %>% 
              dplyr::select(ISO2, lights2013))
```

## Importing data from WDI

Data about night-time lights intensity will be correlated with GDP and population, similarly as in previous examples. For countries of the world such data can be imported in R directrly from the [World Development Indicators](https://datacatalog.worldbank.org/dataset/world-development-indicators) database with the `WDI` package.

To request the indicator GDP (Current US$) we need to use its code [`NY.GDP.MKTP.CD`](http://api.worldbank.org/v2/indicator/NY.GDP.MKTP.CD).

```{r}
GDP_world <- WDI(country = "all", 
                 indicator = "NY.GDP.MKTP.CD",
                 start = 2013,
                 end = 2013)
```

One can also search for the code with the `WDIsearch()` function.

```{r}
WDIsearch(string = "gdp") %>% 
  head(10)
```

We also need data on population:

```{r}
pop_world <- WDI(country = "all", 
                 indicator = "SP.POP.TOTL",
                 start = 2013,
                 end = 2013)
```

Let's put both data together.

```{r}
world_data <- merge(GDP_world, pop_world)

head(world_data)
```

There is a column called `iso2c` which includes the same country codes as `ISO2` in the map file, so data can be easily put together with the map object. Lets give informative names to the last two columns.

```{r}
map_world <- map_world %>% 
  left_join(world_data, by = c("ISO2" = "iso2c")) %>% 
  rename(pop2013 = SP.POP.TOTL,
         gdp2013 = NY.GDP.MKTP.CD)
```

## Correlations

In the last step we will calculate the correlation matrix.

```{r}
map_world %>% 
  mutate(l_lights2013 = log(1+lights2013)) %>% 
  dplyr::select(lights2013, l_lights2013, gdp2013, pop2013) %>% 
  st_drop_geometry() %>% 
  cor(use = "pairwise.complete.obs")
```

and present relationships graphically.

```{r}
ggplot(map_world,
       aes(x = gdp2013/1e+12,
           y = lights2013/1e+7)) +
  xlab("GDP in 2013") +
  ylab("NTLI in 2013") +
  geom_point(size = 5, 
             col = "darkblue") +
  geom_smooth(method = "lm", se = FALSE) +
  geom_label_repel(data = map_world %>% 
                     filter(gdp2013/1e+12 > 3 |
                              lights2013/1e+7 > 1.5),
                   aes(label = country),
                   size = 5) +
  geom_text(x = 5, y = 6.5,
            size = 8,
            label = "cor(GDP, NTLI) = 0.91") +
  geom_text(x = 5, y = 6,
            size = 8,
            label = "cor(GDP, lnNTLI) = 0.39") +
  ggtitle("Night-time lights intensity vs GDP in 2013 in world countries") +
  theme_bw()
```

And the same for population vs night-time lights.

```{r}
ggplot(map_world,
       aes(x = pop2013/1e+8,
           y = lights2013/1e+7)) +
  xlab("population in 2013") +
  ylab("NTLI in 2013") +
  geom_point(size = 5, 
             col = "darkblue") +
  geom_smooth(method = "lm", se = FALSE) +
  geom_label_repel(data = map_world %>% 
                     filter(pop2013/1e+8 > 2.5 |
                              lights2013/1e+7 > 1.5),
                   aes(label = country),
                   size = 5) +
  geom_text(x = 8, y = 6.5,
            size = 8,
            label = "cor(pop, NTLI) = 0.57") +
  geom_text(x = 8, y = 6,
            size = 8,
            label = "cor(pop, lnNTLI) = 0.33") +
  ggtitle("Night-time lights intensity vs population in 2013 in world countries") +
  theme_bw()
```
